---
title: "TO404Proj1"
author: "Jack Zimmerman, Danielle Bidigare, Jeff Sondheimer, John McCarthy, Eliza Brown"
date: "11/22/2020"
output: 
  html_document: 
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

# Introduction

## Packages Used
```{r}
library(tidyverse)
library(ggplot2)
library(geosphere)
library(OneR)
library(lubridate)
library(leaflet)
library(readr)
```

## File Loading
```{r, cache = TRUE}
#Loading a pre-made sample csv file made from another RMD file that took a 5% sample of every data monthly dataset

citibike <- read.csv("citibike_sample_total.csv")
```

## Variable Elimination
```{r, cache = TRUE}
str(citibike)

#Getting rid of the X variable as it is just a counter
citibike$X <- NULL


```
## Data Cleaning

### Factorizing Variables
```{r, cache = TRUE}
#Setting all of the categorical variables to a factor
citibike$start.station.id <- as.factor(citibike$start.station.id)
citibike$start.station.name <- as.factor(citibike$start.station.name)

citibike$end.station.id <- as.factor(citibike$end.station.id)
citibike$end.station.name <- as.factor(citibike$end.station.name)

citibike$bikeid <- as.factor(citibike$bikeid)
citibike$usertype <- as.factor(citibike$usertype)

citibike$gender <- as.factor(citibike$gender)
#Give name to numeric gender classification
levels(citibike$gender)[levels(citibike$gender)=="0"] <- "Unknown"
levels(citibike$gender)[levels(citibike$gender)=="1"] <- "Male"
levels(citibike$gender)[levels(citibike$gender)=="2"] <- "Female"


#Creating an age column that will be easier to intepret
citibike$age <- 2019 - citibike$birth.year
head(citibike$age)

```
### Time and Data Cleaning
```{r, cache = TRUE}



citibike$months <- months(as.Date(citibike$starttime))
citibike$months <- as.factor(citibike$months)

mode(citibike$months)
str(citibike$months)

citibike = citibike %>% mutate(date = substr(citibike$starttime, 1, 10) %>% parse_date(),
                               time = substr(citibike$starttime,12,19) %>% parse_time())

str(citibike$date)
str(citibike$time)

str(citibike)

```
### Age Cleaning
```{r, cache = TRUE}
head(sort(citibike$age, decreasing = TRUE), 10)

#We don't know the reason why the age is wrong (they could have clicked on 1880 instead of 1980, but we can't justify this), so I am going to set the age of anyone over 90 to the mean of the sample

citibike$age <- ifelse(citibike$age >90, NA, citibike$age)
citibike$age <- ifelse(is.na(citibike$age) == TRUE, mean(citibike$age, na.rm = TRUE), citibike$age)

summary(citibike$age)

head(sort(citibike$tripduration, decreasing = TRUE), 10)
#All of these are likely "forgotten" and not returned. I wonder what the credit card bill looked like

filter(citibike, tripduration >1000000)
```
### Distance Between Stations
```{r, cache = TRUE}
#Load geosphere so we can use distHaversine function
library(geosphere)
#
startlocation <- select(citibike,c(start.station.latitude, start.station.longitude))
endlocation <- select(citibike, c(end.station.latitude, end.station.longitude))

str(startlocation)

citibike$distance <- distHaversine(select(citibike,c(start.station.latitude, start.station.longitude)),select(citibike, c(end.station.latitude, end.station.longitude)))

head(citibike$distance)
```


# Data Exploration

## Basic Statistics
```{r, cache = TRUE}
str(citibike)
avergae_trip_duration <- mean(citibike$tripduration)
median_trip_duration <- median(citibike$tripduration)
shortest_trip <- min(citibike$tripduration, na.rm=TRUE)
longest_trip <- max(citibike$tripduration, na.rm=TRUE)
```

The longest trip is `r longest_trip` seconds, and the shortest trip is `r shortest_trip` seconds.

Seeing how many "forgotten" bike trips there are. It's unlikely that anyone would have a bike greater than a week, so that's the criteria we'll use for a forgotten bike


## Forgotten Bikes
```{r, cache = TRUE}
forgotten_bikes <- nrow(citibike[citibike$tripduration > 60*60*24*7,])
forgotten_bikes

#See what percent of forgotten bikes are from customers. 
round(nrow(citibike %>%
  filter( tripduration > 60*60*24*7 & usertype == "Customer") %>%
  select(c(tripduration, usertype)))  / 
  nrow(citibike %>%
  filter( tripduration > 60*60*24*7) %>%
  select(c(tripduration, usertype)))*100, digits = 2)
```

We have `r forgotten_bikes` bike trips in our data sample.


## User Type
```{r, cache = TRUE}
total_number_user <- nrow(citibike[citibike$usertype,])
total_number_user

number_customer <- nrow(citibike[citibike$usertype=="Customer",])
number_subscriber <- nrow(citibike[citibike$usertype=="Subscriber",])
number_customer/total_number_user
number_subscriber/total_number_user
```
## Most Popular Station
```{r, cache = TRUE}
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

getmode(citibike$start.station.name)
#This is right next to the Chrysler building in Manhattan, so Pershing Square makes sense as the most popular start station

#Getting number of trips started at Pershing Square North
nrow(citibike[citibike$start.station.name == "Pershing Square North",])


#Most popular end station. I'm curious to see if Pershing Square is also the most popular end destination
getmode(citibike$end.station.name)
#and it is the most popular


#Representative share of most popular station

#Expected number of trips per station with an equal distribution
nrow(citibike)/length(levels(citibike$start.station.name))

#Pershing Square trips as a percentage of expected number of trips

nrow(citibike[citibike$start.station.name == "Pershing Square North",])/(nrow(citibike)/length(levels(citibike$start.station.name)))
#There is a 7x increase in the expected number of trips

```

## More data exploration
```{r, cache = TRUE}
#Avg trip duration by user
avg_tripduration_by_user <- round(tapply(citibike$tripduration,citibike$usertype,mean,na.rm=TRUE),2)
avg_tripduration_by_user
tripduration_baseplot <- ggplot(data=citibike,aes(x=usertype,y=tripduration))
tripduration_baseplot+geom_boxplot(outlier.colour="black",outlier.shape=16,outlier.size=2,notch=TRUE)+coord_cartesian(ylim=c(0,2500))


```


# User Demographic Data Visualization

## Trip Duration Density by User Type

```{r, cache = TRUE}

#density of user types by trip duration
ggplot(data = subset(citibike, tripduration<24*60), aes(x=tripduration, fill=usertype)) + geom_density(alpha=.5)

```
## Time of Rides by User Type

```{r, cache = TRUE}
## Notes

#density of usertype by time

citibike = citibike %>% mutate(date = substr(citibike$starttime, 1, 10) %>% parse_date(),
                               time = substr(citibike$starttime,12,19) %>% parse_time())


ggplot(data = citibike, aes(x = time, fill = usertype)) + geom_density(alpha=.5) + scale_fill_discrete("User Type")

#stats for subscribers trip duration at 6AM to 9AM and 4PM to 7PM (ie rush hour times)
citibike %>% filter(time>parse_time("6:00:00") & time<parse_time("9:00:00") & usertype == "Subscriber") %>% select(tripduration) %>% summary()

citibike %>% filter(time>parse_time("16:00:00") & time<parse_time("19:00:00") & usertype == "Subscriber") %>% select(tripduration) %>% summary()


citibike %>% group_by(date) %>% count()

```

## Number of Trips by Months 

```{r, cache = TRUE}
citibike$months <- factor(citibike$months, levels = c("January","February", "March","April", "May", "June", "July", "August", "September", "October", "November", "December"))
ggplot(data = citibike, aes(x= months, fill = gender)) + geom_bar() + labs(title = "Number of Trips by Month", x = "Month", y = "Number of Trips")
ggplot(data = citibike, aes(x= months, fill = usertype)) + geom_bar() + labs(title = "Number of Trips by Month", x = "Month", y = "Number of Trips")
```

# User Speed

## Calculating Speed
```{r, cache = TRUE}


citibike$distance <- distHaversine(cbind(citibike$start.station.longitude, citibike$start.station.latitude), cbind(citibike$end.station.longitude, citibike$end.station.latitude))
summary(citibike$distance)

#Create speed column and convert it to MPH by multiplying by 2.237
citibike$speed <- (citibike$distance / citibike$tripduration) * 2.237

summary(citibike$speed)

#bin the time values and create ggplot - x value speed (seconds on x axis and speed on y axis - smooth graph)

bin(citibike$time, nbins = 24, labels = NULL, method = c("length", "content",
  "clusters"), na.omit = TRUE)


#pull out time—mutate or group_by?, group by time, find average speed for each group

citibike$hourbin <- as.factor(hms(citibike$time)$hour)
```

## Speed and Time Relationship

```{r}

ggplot(data=citibike,aes(x=hourbin,y=speed,group=hourbin))+geom_boxplot()+coord_cartesian(ylim=c(0, 23))

ggplot(data=citibike,aes(x=hourbin,y=speed,fill=usertype))+geom_boxplot()+coord_cartesian(ylim=c(0, 23))

ggplot(data=citibike,aes(x=hourbin,y=speed,fill=usertype))+geom_boxplot()+coord_cartesian(ylim=c(0, 23))+facet_wrap(citibike$weekday)

##Look at speed by weekday

citibike$weekday <- as.factor(wday(citibike$date,label=TRUE))
levels(citibike$weekday)

ggplot(data=citibike,aes(x=weekday,y=speed,fill=usertype))+geom_boxplot()+coord_cartesian(ylim=c(0, 20))

##Look at speed by month

 ggplot(data=citibike,aes(x=months,y=speed,fill=usertype))+geom_boxplot()+coord_cartesian(ylim=c(0, 20))

#Plot mean speed by hour
 
avgspeed_hour <- round(tapply(citibike$speed,citibike$hourbin,mean),2)

plot(avgspeed_hour,pch=15,col="red")
```


# Station Deficit

## Calculating Station Deficit

```{r, cache = TRUE}
#bike departures and bike arrivals 
str(citibike)

bike_departures <- group_by(citibike, station = citibike$start.station.name, latitude = citibike$start.station.latitude, 
                            longitude = citibike$start.station.longitude) #should bike id be as a factor or number 
departure_count <- dplyr::summarise(bike_departures, departure_count = n()) 

bike_arrivals <- group_by(citibike, station = citibike$end.station.name, latitude = citibike$end.station.latitude, 
                            longitude = citibike$end.station.longitude)
arrival_count <- dplyr::summarise(bike_arrivals, arrival_count = n()) #ask evan the reason we need to make it into days 

#merge departure and arrival data 
bike_deficit <- merge(departure_count, arrival_count, all = TRUE)

#make Na's = 0 
bike_deficit[is.na(bike_deficit)] <- 0

#make deficit variable == departures - arrivals
bike_deficit$deficit <- bike_deficit$departure_count - bike_deficit$arrival_count
```

## Station Mapping

### Deficit Map
```{r}
leaflet(bike_deficit) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = bike_deficit$longitude, lat = bike_deficit$latitude,
                   popup = paste(bike_deficit$station, "<br>", ifelse(bike_deficit$deficit>=0, "Bike deficit = ", "Bike surplus = "), 
                               abs(bike_deficit$deficit)  ), 
                   radius = abs(bike_deficit$deficit)/500, color = ifelse(bike_deficit$deficit>0, "red", "green"))

#create leaflet/gganimation for top 5 locations - day of the week
```

### Top Stations Map
```{r}
#Mapping the top 5 start and end stations




table(citibike$end.station.id) %>% sort(decreasing = T) %>% head(5)

#plotting the most frequent start stations 

top_5start <- citibike %>% group_by(start.station.name, start.station.latitude, start.station.longitude) %>% summarise(count = n()) %>% arrange(desc(count))
top_5start <- head(top_5start,5) 
str(top_5start)


leaflet(top_5start)%>%
  addTiles()%>%
  addCircleMarkers(lng=top_5start$start.station.longitude,lat=top_5start$start.station.latitude,
                   popup=paste(top_5start$start.station.name,"<br>","Ride launches per year",top_5start$count),
                   radius=top_5start$count/300,color="green")


top_5end <- citibike %>% group_by(end.station.name, end.station.latitude, end.station.longitude) %>% summarise(count = n()) %>% arrange(desc(count))
top_5end <- head(top_5end,5)
str(top_5end)

##ideas: bin by month (possibly do animation), color map by time (would need new dataframe of starttime)




leaflet()%>%
  addTiles()%>%
  addCircleMarkers(lng=top_5end$end.station.longitude,lat=top_5end$end.station.latitude,
                   popup=paste(top_5end$end.station.name,"<br>","Ride ends per year:",top_5end$count),
                   radius=top_5end$count/300,color="red")%>%
  addCircleMarkers(lng=top_5start$start.station.longitude,lat=top_5start$start.station.latitude,
                   popup=paste(top_5start$start.station.name,"<br>","Ride launches per year:",top_5start$count),
                   radius=top_5start$count/300,color="blue")


#plot on leaflet

nrow(bike_deficit)
summary(citibike)
```


# Weather Impacts

## Itroducing Weather Data

### File Loading
```{r, cache = TRUE}
#This data  is taken from NOAA website
tempdata <- read.csv("NYCWeather2019.csv")


```

### Data Cleaning
```{r}

str(tempdata)
summary(tempdata)

table(tempdata$STATION)
#All the same station, so we will get rid of this information. This station is located in Central Park, which is a great central point and fairly representative of the entire city
tempdata$STATION <- NULL
tempdata$NAME <- NULL


summary(tempdata$TAVG)
#Conventional average temperature is calculated by summing the maximum and minimum temperatures and dividing by two. That's what will be done for TAVG here
tempdata$TAVG <- (tempdata$TMIN + tempdata$TMAX)/2


#Doing some basic searches, AWND means average daily wind speed in meters per second, and SNWD means snow depth in millimeters. Snow depth implies how much snow is actively on the ground whereas snow implies how much fell from the sky (melting impacts and residual snow from days before)

#For the one NA in snow, it's fair to assume that it didn't snow that day.
tempdata$SNOW <- ifelse(is.na(tempdata$SNOW) == TRUE, 0, tempdata$SNOW)
```


### Data Frame Merge 
```{r, cache = TRUE}


tempdata$DATE <- as.Date(tempdata$DATE, format = "%m/%d/%Y")
citibike <- merge(citibike, tempdata, by.x = "date", by.y = "DATE",no.dups = TRUE)
str(citibike)
```

### Weather List Data Creation 
```{r, cache = TRUE}
#Creating variables that are grouped by month and day level to analyze trends
avg_month_temp <- citibike %>% group_by(months) %>% summarise(mean(TAVG), mean(TMIN), mean(TMAX),count = n())
avg_month_temp <- avg_month_temp %>% rename(tavg = "mean(TAVG)",
                                            tmin = "mean(TMIN)",
                                            tmax = "mean(TMAX)")



day_weather_data <- citibike %>% group_by(date) %>% summarize(mean(TAVG), mean(AWND), mean(SNOW), mean(SNWD), mean(tripduration), count = n())

day_weather_data <- day_weather_data %>% rename(tavg = "mean(TAVG)",
                                          tripduration = "mean(tripduration)",
                                          awnd  = "mean(AWND)",
                                          snow = "mean(SNOW)",
                                          snow_depth = "mean(SNWD)")
```

## Weather Exploratiom

```{r}


ggplot(data = avg_month_temp, aes(x = months, y = tavg)) +geom_point() + geom_pointrange(aes(ymax = tmax, ymin = tmin)) +theme(axis.text.x = element_text(angle = 90))

#(data = avg_month_temp, aes(x = months)) + geom_bar(stat = "identity", aes(y = count)) + geom_pointrange(aes(y = tavg, ymax = tmax, ymin = tmin)) + scale_y_continuous(name = "Number of Rides", sec.axis = sec_axis(trans = ~10, name = "Average Temperature (F)"))

ggplot(data = day_weather_data, aes(x = tavg, y = awnd)) + geom_point() + geom_smooth()

str(citibike)
```


## Optimal Riding Temperature 
```{r, cache = TRUE}

ggplot(data = avg_month_temp, aes(x = months, y = tavg)) +geom_point() + geom_pointrange(aes(ymax = tmax, ymin = tmin)) +theme(axis.text.x = element_text(angle = 90))


ggplot(data = day_weather_data,aes(x = tavg, y = count)) + geom_point() + geom_smooth()

user_weather_chart <-  ggplot() + geom_point(data =(citibike %>% group_by(date) %>% filter(usertype == "Subscriber") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "red")) + geom_smooth(data = (citibike %>% group_by(date) %>% filter(usertype == "Subscriber") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")),aes(x = tavg, y = count)) + geom_point(data = (citibike %>% group_by(date) %>% filter(usertype == "Customer") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "green")) + geom_smooth(data = (citibike %>% group_by(date) %>% filter(usertype == "Customer") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "green"))

user_weather_chart

age_weather_chart <-  ggplot() + geom_point(data =(citibike %>% group_by(date) %>% filter(age > 65) %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "red")) + geom_smooth(data = (citibike %>% group_by(date) %>% filter(age > 65) %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")),aes(x = tavg, y = count)) + geom_point(data = (citibike %>% group_by(date) %>% filter(age <30) %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "green")) + geom_smooth(data = (citibike %>% group_by(date) %>% filter(age < 30) %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "green"))

age_weather_chart

gender_weather_chart <- age_weather_chart <-  ggplot() + geom_point(data =(citibike %>% group_by(date) %>% filter(gender == "Male") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "red")) + geom_smooth(data = (citibike %>% group_by(date) %>% filter(gender == "Male") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")),aes(x = tavg, y = count)) + geom_point(data = (citibike %>% group_by(date) %>% filter(gender == "Female") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "green")) + geom_smooth(data = (citibike %>% group_by(date) %>% filter(gender == "Female") %>% summarize(mean(TAVG),count = n()) %>%
  rename(tavg = "mean(TAVG)")), aes(x = tavg, y = count, color = "green"))

gender_weather_chart

```

## Wind Speed Impacts
```{r, cache = TRUE}
ggplot(data = day_weather_data,aes(x = awnd, y = count)) + geom_point(aes(colour = (tavg < 35))) + geom_smooth()

ggplot(data = day_weather_data,aes(x = awnd, y = tripduration)) + geom_point(aes(colour = (tavg < 35))) + geom_smooth()
#Affects numbers of trips, windspeed above 6, a lot less "long distance" rides
```


## Snow Impacts
```{r, cache = TRUE}
#Filtering out for the days in which it snowed in NYC
citibike %>% group_by(date) %>% filter(SNOW >0) %>% summarize(count = n(), mean(TAVG), mean(tripduration))
#Looks like it snowed 14 days in NYC


ggplot(data = day_weather_data, aes(x = tavg, y = count, colour = (snow > 0))) + geom_point()
ggplot(data = day_weather_data, aes(x = date, y = tavg, colour = (snow_depth > 0))) + geom_point()
ggplot(data = day_weather_data, aes(x = date, y = tripduration, colour = (snow > 0))) + geom_point()
```

# Summary

## Findings Summary

## Recommendation